---
title: "A first encounter with data: graphs"
author: "Nathan Brouwer"
date: "June 28, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Preliminaries

#### Load packagtes
```{r}
library(wildlifeR)
library(ggplot2)
library(cowplot)
library(ggpubr)
library(dplyr)
```

#### Load data
```{r}
data(frogarms)
```


#### Subset your data

The function make_my_data2L() will extact out a random subset of the data.  Change "my.code" to your school email address, minus the "@pitt.edu" or whatever your affiliation is.
```{r}
my.frogs <- make_my_data2L(dat = frogarms, 
                           my.code = "nlb24", # <=  change this!
                           cat.var = "sex",
                           n.sample = 20, 
                           with.rep = FALSE)

```



### Boxplots

Basic boxplot
```{r}
ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex")
```



Notched boxplot.


We'll use the original frogarms dataframe first for this
THese aren't commonly used; the notches work kind of like confidence intervals to determine if medians are different.  
```{r}
ggboxplot(data = frogarms,
          y = "mass",
          x = "sex",
          notch  = TRUE)
```

Now try your own subset of the data.  The Notch calculations likely get messed up with small samples sizes)
```{r}
ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex",
          notch  = TRUE)
```



Add colored fill; note that it is "fill" not "color".  Color changes the color of the lines
```{r}
ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex",
          notch  = TRUE,
          fill = "sex")
```

We can turn off the notchign by adding a "#" character before it.  This is called "commenting out"

```{r}
ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex",
          #notch  = TRUE,
          fill = "sex")
```


Add raw data.    This works best with small datasts
```{r}
ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex",
          #notch  = TRUE,
          fill = "sex",
          add = "point")
```


Jiter raw data
This can be helpfu, though ggpubr::ggboxplot doesn't allow much control over the "jittering".  Jittering helpful when you have large datsets and want to avoid overlap in the points.
```{r}
ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex",
          #notch  = TRUE,
          fill = "sex",
          add = "jitter")
```



Label axes
```{r}
ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex",
          notch  = TRUE,
          fill = "sex",
          add = "jitter",
          xlab = "Sex",      #x axis (horizontal)
          ylab = "Mass (g)") #y axis (vertical)
```

Add title
not usually done for publication but useful for keeping track of things and for presentations
```{r}
ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex",
          notch  = TRUE,
          fill = "sex",
          add = "jitter",
          xlab = "Sex",
          ylab = "Mass (g)",
          main = "Mass of Australian frogs by sex") #Main title
```



Move legend to bottom
```{r}
ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex",
          notch  = TRUE,
          fill = "sex",
          add = "jitter",
          xlab = "Sex",
          ylab = "Mass (g)",
          main = "Mass of frogs by sex",
          legend = "bottom")
```


Change color pallete
```{r}
ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex",
         # notch  = TRUE,
          fill = "sex",
          add = "jitter",
          xlab = "Sex",
          ylab = "Mass (g)",
          main = "Mass of frogs by sex",
          legend = "bottom",
          palette = c("green","blue"))
```



#### Plotting multple plots with cowplot::plot_grid

We can save a plot to an R object
```{r}
gg.my.frogs <- ggboxplot(data = my.frogs,
          y = "mass",
          x = "sex")


```


Call just the object (eg, just type it into the console. or highlight jsut the word)
```{r}
gg.my.frogs
```

Make an object using the frogarms data
```{r}
gg.frogarms <- ggboxplot(data = frogarms, #use original data
          y = "mass",
          x = "sex")

```


Now plot both

```{r}
plot_grid(gg.my.frogs,
          gg.frogarms)
```



Add labels.  Note that alignment is off sometimes.

```{r}
plot_grid(gg.my.frogs, 
          gg.frogarms,
          labels = c("a)My fogs","b)All the frogs"))
```





### Plot means with error bars

Super hand function ggerrorplot()
Default is mean +/- 1 standard error
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex")
```



Mean and se; note sd is not often used in publications
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_sd")
```


Mean and 95% confidence interval
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci")
```


Plot your data and original data
```{r}
#your data
gg.my.frogs <- ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci")

#all of the data
gg.all.frogs <- ggerrorplot(data = frogarms, #change data
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci")

plot_grid(gg.my.frogs, gg.all.frogs)
```

Set colors
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex")
```


Add raw data.  kinda crazy
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex",
          shape = "sex",
          add = "point")
```


Jitter raw data.  even crazier
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex",
          add = "jitter")
```



Back to just the means Increase size
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex",
          size = 1.5) #
```



Move legend to the bottom

?set all of this stuff as eval = F and have students figur eout how to add it?
m aybe not - goal is just to do "1st encounter"
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex",
          size = 1.5,
          xlab = "Sex",
          ylab = "Mass (g)",
          legend = "bottom") #
```






## arms
```{r}
ggerrorplot(data = frogarms,
          y = "arm",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex",
          size = 1.5,
          xlab = "Sex",
          ylab = "Mass (g)",
          legend = "bottom") #
```


