---
title: "Untitled"
author: "Nathan Brouwer"
date: "June 28, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



### Preliminaries

#### Load packagtes
```{r}
library(wildlifeR)
library(ggplot2)
library(cowplot)
library(ggpubr)
library(dplyr)
```

#### Load data
```{r}
data(frogarms)
```


#### Subset your data

The function make_my_data2L() will extact out a random subset of the data.  Change "my.code" to your school email address, minus the "@pitt.edu" or whatever your affiliation is.
```{r}
my.frogs <- make_my_data2L(dat = frogarms, 
                           my.code = "nlb24", # <=  change this!
                           cat.var = "sex",
                           n.sample = 20, 
                           with.rep = FALSE)

```



t-test used to tell if two groups are different


### T-test
```{r echo  = F}
t.test(mass ~ sex, data = my.frogs)
```
spits out R's standard t.test table

Save to object
```{r}
mass.t <-  t.test(mass ~ sex, data = my.frogs)
```



-----------------

look at w/broom::glance.  relables thiugns a bit odd and would be nice to round.  will stick with original R output
```{r}
library(broom)
glance(mass.t)
```
------------------


```{r}
mass.t
```


Check the means using dplyr
```{r}
my.frogs %>% group_by(sex) %>% summarize(mean.mass = mean(mass))
```


What does all of this mean?

Quiz

p = 
p interpretation = 
df = 
why df fractional? (this one is hard!)
t = 
What would happen to p if t was bigger?
What is a "95% CI"
What is this a 95% CI for?



What does the CI mean?

```{r}
plot_t_test_ES(mass.t)
```




Make a plot of the means with error bars.  Save to an object called gg.means

```{r}
gg.means <-ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci") +
  ggtitle("Group means & error bars")
```

Save effect size

```{r}
gg.ES <-plot_t_test_ES(mass.t) +
  ggtitle("__________ & errorbars")
```


Plot both
```{r}
plot_grid(gg.means,gg.ES)
```




A hint
```{r}
gg.means <-ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci") +
  ylab("Mass (g)") +
  xlab("Sex") +
  ggtitle("Group means & error bars")


gg.ES <-plot_t_test_ES(mass.t) +
  ylab("Difference between groups (g)") +
  ggtitle("__________ & errorbars")
```

```{r}
plot_grid(gg.means,gg.ES)
```




Did anyone get a significant result?



### Arm girth

(do same thing for arm girth.  see the super significant values?)



Now we are going to unpack this




------------------------------

There is a **major** flaw in this analysis.  consider the following graph where the mass is plotted on the x-axis and the arm girth is plottedon the y-axis.  Does arm girth vary just because of sex, or becasuse of sex and mass?

This is ANCOVA.  Sometimes taught as an extension of ANOVA, or as a type of regression.  

```{r}
ggscatter(data = frogarms,
          y = "arm",
          x = "mass",
          color = "sex")
```

------------------------------
