---
title: "Intro to 1-way ANOVA: impacts of diet on deer antlers"
author: "Nathan Brouwer  brouwern@gmail.com  @lobrowR"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Merging w/ dplyr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE)
```


## Introduction

The purpose of this lab is to learn the basics of 1-way ANOVA in R.  We will explore data where white-tailed deer where fed different diets over the course of the spring and summer of one year.

* Diet 1: "Hi.Hi" = deer fed high protein diets during spring and summer
* Diet 2: "Hi.Lo" = deer fed high quality diets during spring, but low quality diet during summer
* Diet 3: "Lo.Hi" = deer fed low-protein diet during spring but high quality diet during summer.

Note: there is no "Lo.Lo" treatment

Analyis with 1-way ANOVA will allow us to determine whether

1. All of the means are similar or if at least one differs
2. Which treatment, if any, is most likely impacting antler growth.


### References
Data were simulated from summary parameters in:

Asleson et al. 1997.  Effects of seasonal protein restriction on antlerogenesis and body mass in adult male white-tailed deer.  Journal of Wildlife Management 61.

For more information see ?antlers



## Part 1: Summary stats & plotting means

### 1) Summary stats on all data

In this section the grand mean on ALL the data is calculated; data is NOT broken up by treatment!  Data is considered by subgroup below

Load data

```{r}
library(wildlifeR)
data(antlers)
```


```{r}
# total sample size (all observations)
dim(antlers)

n.total <- length(antlers$mass)

#mean of ALL samples
summary(antlers)

mean(antlers$mass)

#variance of ALL samples
var(antlers$mass)

#stdev of ALL samples
mass.sd <- sd(antlers$mass)
```



### 2) Standard error for all  data

This ignores treatments.  All data are combined / pooled.

We'll do this a couple different ways to show how R code can vary.


Broken up into steps, saving an object at each stuep
```{r}
#square root of N
sqrt.n <- sqrt(n.total)

#the see
mass.se <- mass.sd/sqrt.n

mass.se

```

Doing it a little "on fly" 
```{r}

mass.se <- mass.sd/sqrt(n.total)


```



Totally on the fly - nothing pre-saved as an object
```{r}

#Using raw data
mass.se <- sd(antlers$mass)/
             sqrt(length(antlers$mass))
```



### 3) 95% CI for all data

This is the 95% confidence interval around the overall/grand mean.  W approximate the CI as 1.96*SE

```{r}
1.96*mass.se
```


### 4) Calcualte summary stats for each group

* Calculae the mean for each diet treatment using dplyr.  
* THe group_by() function split it up by treatment
* summarize() calcualtes the mean
* Recall the "%>%" is called a "pipe"

See pages 70-73 in "Getting started with R: An INtroduction for Biologists" for information on summarize() and group_by().  In particular see section 3.7.3 on page 72: "Method 2: Pipe, no nesting"

```{r}
library(dplyr)

antlers %>% group_by(diet) %>%
   summarize(mass.mean = mean(mass))

```

We can get the SD like this
```{r}

antlers %>% group_by(diet) %>%
   summarize(mass.sd = sd(mass))

```


And the sample size like this
```{r}

antlers %>% group_by(diet) %>%
   summarize(mass.N = length(mass))

```



We can calcualte all of them like below.  this time we'll store the ouptut in an object "mass.means"
```{r}

mass.means <- antlers %>% group_by(diet) %>%
   summarize(mass.mean = mean(mass),
             mass.sd = sd(mass),
             mass.N = length(mass))

```



### 5) Calcualte the 3 SE values

```{r}

mass.SEs <- mass.means$mass.sd/sqrt(mass.means$mass.N)


```

Add them to the dataframme


```{r}

mass.means$mass.SEs <- mass.SEs

```


We can conver these to approximately 95% confidence intervals like this

```{r}
mass.means$mass.CI95 <- mass.means$mass.SEs*1.96
```


What do you notice about the relationship between the SD, SE, and 95% CI?


## 6) Plot the 3 mean values

We'll plot means using ggpubr.  We'll give the function ggerrorplot() the raw data and it will calculate the means etc to make the plot.

See the link below for more information
http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/79-plot-meansmedians-and-error-bars/

mean_se
mean_sd
mean_range

First, we'll plot the means and their standard deviations SD
```{r}
library(ggpubr)

ggerrorplot(antlers, 
            x = "diet", 
            y = "mass", 
            desc_stat = "mean_sd",
            add = "mean"
            )

```


Next we'll plot the standard errors.  What happens to the x-axis?

```{r}
ggerrorplot(antlers, 
            x = "diet", 
            y = "mass", 
            desc_stat = "mean_se",
            add = "mean"
            )

```



We can make a plot with both if we save the plots to R objects.    We'll use the argument ylim = c(400,1000) to make it so they both have the same y axes

Note: We'll need to load the gridExtra package

```{r}
library("gridExtra")
plot1 <- ggerrorplot(antlers, 
            x = "diet", 
            y = "mass", 
            desc_stat = "mean_sd",
            add = "mean",
            ylim = c(400,900) #set axes
            )

plot2 <- ggerrorplot(antlers, 
            x = "diet", 
            y = "mass", 
            desc_stat = "mean_se",
            add = "mean",
            ylim = c(400,900) #set axes
            )

```


Plot both plots.  What do you notice about the standard errors?
```{r}
grid.arrange(plot1,plot2)

```


We can also easily make a plot using confience intervals.
```{r}
ggerrorplot(antlers, 
            x = "diet", 
            y = "mass", 
            desc_stat = "mean_ci",
            add = "mean",
            ylim = c(400,900) #set axes
            )
```


**We have not calculated any p-values yet, but can you use the "inference by eye" approach to judge whether any of the means are likely to be "significantly" different based on the 95% confidence intervals?**



## Part 2: Omnibus ANOVA test

An "omnibus ANOVA test" (or "omnibus F test") is used to gauge whether all the means are approximatley equal or whether at least one is likely to be different than the others.  "Omnibus" means "overall".

### 7 & 8) Build the null (Ho) and alternative (Ha) models

We build a null modle represent the null hypothesis that there is no difference between the any of the groups.  Not that we code "mass ~ 1" to represent this null model, which has just a single mean value for all of the data.

```{r}

model.null <- lm(mass ~ 1, 
                 data = antlers)

```


We buidl an alternative model or "model of interest" to test the hypothesis we are interested in: that diet impacts antler growth in some way.  This model will have 3 means: one for each treatment.

```{r}


model.alt <- lm(mass ~ diet, 
                 data = antlers)
```



## 9-12) Conduct the Omnibus test and get the ANOVA output

Produce the ANOVA table using the anova() command.  This will give us the F statistics (the test statistics for ANOVA), degrees of freedom, and our p-values.  (It will also give us some other info that we'll ignore for now but that is typically reported when you do an ANOAV)

```{r}
anova(model.null,
      model.alt)
```






### OPTIONAL:

If you are curious and things are going well, do the optional activities.  These are completely optional



#### Optional: The summary command on an lm() object

Look at the output and how it compares to what we did with anova() on the two models.

```{r}
summary(model.null)
summary(model.alt)

```


#### Optional:  The anova command on a single lm() object

Look at the output and how it compares to what we did with anova() on the two models.

```{r}
anova(model.alt)
```


#### Optional: Change how model is defined

We'll add a "-1" to the model statemnt and call summary().  How does this differe from summary(model.alt)?  (The difference is subtle...)

```{r}
model.alt.2  <- lm(mass ~-1 + diet, data = antlers)

summary(model.alt.2)
```


#### Optional: Plotting model diagnostics

We can plot the "model diagnostics" using plot().  We will go into further details on this topic in the future.

```{r}
plot(model.alt)
```


## Part 3: Pairwise comparisons

### 13-15) Get pairwise p-valeus
```{r}
pairwise.t.test(x = antlers$mass,
                g = antlers$diet,
                p.adjust.method = "none")
```




## Part 4: Pairwise comparisons, correcting for multiple comparisons, & effect sizes


### 17) Build model with aov() function
```{r}

model.alt.aov <- aov(mass ~ diet, 
                     data = antlers)


```

### 18) Get p-values with Tukey's HSD 

```{r}

TukeyHSD(model.alt.aov)

```




### 19) Plot effect sizes

```{r}

tukey.out <- TukeyHSD(model.alt.aov)

plotTukeysHSD(tukey.out)
abline(h = 0, col = 2, lty = 2)

```

