---
title: "Inference-by-eye using RA Fisher's Cat Data"
author: "Nathan Brouwer  brouwern@gmail.com  @lobrowR"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Merging w/ dplyr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE,fig.height = 3, fig.width = 5)
```

A an example of doing "inference by eye" using R.A. Fischer's cat data.  A good summary of this idea is Cummings et al. 2007.  Error bars in experimental biology. http://jcb.rupress.org/content/177/1/7.short


## Load the data
```{r}
library(MASS)
data("cats")
```


## Look at the data
```{r}
dim(cats)
summary(cats)
```


## Summarize the body weight (Bwt) data old-school using summaryBy()

A more modern way would be to use dplyr()
```{r}
library(doBy)

#get the mean and SD
cat.df1 <- summaryBy(Bwt ~ Sex, data = cats, FUN = c(mean,sd))

#get the sample size using length()
cat.df2 <- summaryBy(Bwt ~ Sex, data = cats, FUN = c(length))

#make a combined dataframe
cat.df3 <- merge(cat.df1,cat.df2)

#calculate the standard error SE by hand

cat.df3$SE <- cat.df3$Bwt.sd/sqrt(cat.df3$Bwt.length)

```


Look at the results
```{r}

cat.df3

```


## Plot the data 

### Visualize the raw data

```{r}
par(mfrow = c(1,1),mar = c(3,3.5,1,1))

boxplot(Bwt ~ Sex, data = cats)
```


### Plot the means with error bars

This uses the errbar() function.  A modern contemporary way would use ggplot2 and possibly its extension using ggpubr.

#### The real data 

This is the actual data.  The 95% confidence intervals do not overlap, which indicates that the p-value for the t-test will be less than 0.05.

```{r}
library(Hmisc)
par(mfrow = c(1,2),mar = c(3,3.5,1,1))

y.lim <- c(2.295,3)
errbar(1:2,
       y = cat.df3$Bwt.mean,
       yplus =cat.df3$Bwt.mean + cat.df3$SE,
       yminus = cat.df3$Bwt.mean-cat.df3$SE,
       xlab = "",
       ylab = "",
       xlim=c(0.5,2.5),
       ylim = y.lim,
       xaxt="n",cex =1)
axis(side=1,at=1:2,labels=cat.df3$Sex)
mtext("Sex", side = 1, line = 2, cex = 2)
mtext("Mass (g)", side = 2, line = 2.1, cex = 1.3)

errbar(1:2,
       y = cat.df3$Bwt.mean,
       yplus =cat.df3$Bwt.mean + 1.96*cat.df3$SE,
       yminus = cat.df3$Bwt.mean-1.96*cat.df3$SE,
       xlab = "",
       ylab = "",
       xlim=c(0.5,2.5),
       ylim = y.lim,
       xaxt="n",cex =1)
axis(side=1,at=1:2,labels=cat.df3$Sex)
mtext("Sex", side = 1, line = 2, cex = 2)
mtext("Mass (g)", side = 2, line = 2.1, cex = 1.3)


```


#### Modified data with a non-significant different

Make an alternative version of the data where there isn't a difference between the male and female cats

```{r}
cat.df3.mod <- cat.df3
cat.df3.mod$Bwt.mean[1] <- cat.df3$Bwt.mean[2]-cat.df3$Bwt.mean[2]*0.0425
```


The overlap of the error bars here is greater than 1/2 the length of the bar; therefore the p-value for a t-test will be > 0.05.
```{r}
y.lim <- c(2.6975,3)
par(mar = c(3,3.5,1,1))
errbar(1:2,
       y = cat.df3.mod$Bwt.mean,
       yplus =cat.df3.mod$Bwt.mean + cat.df3.mod$SE,
       yminus = cat.df3.mod$Bwt.mean-cat.df3.mod$SE,
       xlab = "",
       ylab = "",
       xlim=c(0.5,2.5),
       ylim = y.lim,
       xaxt="n",cex =1)
axis(side=1,at=1:2,labels=cat.df3.mod$Sex)
mtext("Sex", side = 1, line = 2, cex = 2)
mtext("Mass (g)", side = 2, line = 2.1, cex = 1.3)

errbar(1:2,
       y = cat.df3.mod$Bwt.mean,
       yplus =cat.df3.mod$Bwt.mean + 1.96*cat.df3.mod$SE,
       yminus = cat.df3.mod$Bwt.mean-1.96*cat.df3.mod$SE,
       xlab = "",
       ylab = "",
       xlim=c(0.5,2.5),
       ylim = y.lim,
       xaxt="n",cex =1)
axis(side=1,at=1:2,labels=cat.df3.mod$Sex)
mtext("Sex", side = 1, line = 2, cex = 2)
mtext("Mass (g)", side = 2, line = 2.1, cex = 1.3)



```



## T-test

A t-test for the differenec between female and male cats.

```{r}

t.test(Bwt ~ Sex, data = cats)


```




```{r}

summary(lm(Bwt ~ -1+Sex, data = cats))

```

















