---
title: "Reporting statistical result from a 2-sample t-test"
author: "Nathan Brouwer  brouwern@gmail.com  @lobrowR"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
html_document: default
vignette: >
  %\VignetteIndexEntry{Reporting t.test}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

## Introduction

This vignette shows an example of reporting the results from a 2-sample t-test using data on the impact of invasive trout on salmon survival.  The data are originaly from Levin et al (2002) are used in an example in chapter 12 of Whitlock & Schulter 2nd.  See ?brook_trout_ABD for more details.

### References
Levin et al.  2002.  Non–indigenous brook trout and the demise of Pacific salmon: a forgotten threat?  PRSB 269.  DOI: 10.1098/rspb.2002.2063

## Outline of tasks

* Load the data into R
* Create a boxplot of the raw data
* State the relevant statistical null (Ho) and alternative (Ha) hypotheses
* Carry out an appropriate t-test
* Report the appropriate results in a full sentence as it would appear in a report or scientific paper


## Dataset up

The data are availabe in the wildlifeR package and can be loaded using data(wildlifeR).  Note that if you use the dataframe in wildlifeR you have to calculate the survival rate by hand.  I will remake the data by hand as an example of making a simple dataframe.

```{r, include=FALSE}
library(wildlifeR)
data(brook_trout_ABD)

#calculate survival
brook_trout_ABD$survival <- brook_trout_ABD$salmon.released/(brook_trout_ABD$salmon.surv+brook_trout_ABD$salmon.released)

#make column of fake data
fake.surv <- brook_trout_ABD$survival
fake.surv[1:6] <- with(brook_trout_ABD,
                       survival - survival*0.15)[1:6]

salmon <- brook_trout_ABD
```

The following code contains the essential parts of the dataframe: a column for the survival rate and for whether brook trout are present or basent.
```{r}
salmon <- data.frame(survival = c(0.83,0.87,0.82,
                                  0.84,0.81,0.84,
                                  0.72, 0.84,0.75,
                                  0.79,0.89,0.87),
                     brook.trout.PRES.ABS =
                       c("present","present","present",            
                         "present","present","present",
                          "absent","absent","absent",
                          "absent","absent","absent"))

```

## Plot raw data

I'll make a boxplot of the raw data using the ggboxplot() function from the package ggpubr, which contains wrappers that extend ggplot2.  Be sure to download and install these packages if needed.

```{r}
#library(ggplot2)
library(ggpubr)

ggboxplot(data = salmon,
          y = "survival",
          x = "brook.trout.PRES.ABS",
          fill = "brook.trout.PRES.ABS",
          xlab = "Brook trout - absent or present?",
          ylab = "Chinook salmon survival")


```


## The hypotheses

The null (Ho) and alternative hypotheses (Ha) are as follows:

* Ho: The survival rates of Chinook salmon are the same whether brook trout are present or absent
* Ha: The presence of brook trout changes survival rates of salmon.

## Do t-test

```{r}
t.test(survival ~ brook.trout.PRES.ABS, 
       data = salmon)
```


## Report the results

For the real data, the results could be reported like this:
"There was no evidence that the mean survival of salmon when brook trout are present (mean = 0.81) is different than when brook trout are absent (mean = 0.84; 2-sample t-test: p = 0.44, t = 0.82, n = 12 streams, df = 6)."

Normally I would also report the standard errors (SE) around the means, but for this exericise we will ignore it.


### Alternative results

What if the results really looked like this?
```{r, echo=FALSE}
t.test(fake.surv ~ salmon$brook.trout.PRES.ABS)
```

The results could be reported like this:
"Surival of chinook salmon in streams where brook were present (mean = 0.71) was significantly lower than when brook trout were absent (mean = 0.81) with a mean difference of 0.10 (95% CI: 0.03-0.17; 2-sample t-test p = 0.012, t = 3.6, n = 12 stream, df = 5.73)"



