---
title: 'ENS 495 Fall 2017 Independent Data Analysis: ANOVA Example Paper'
output:
  pdf_document: default
  word_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


```{r echo = FALSE, warnings = FALSE, message = FALSE}
##source analysis script
library(wildlifeR)
 source("Brouwer_NL_analysis_script_PIWO_00_05_10.R", encoding = 'UTF-8')
```

## TITLE: ... Pileated Woodpecker (Dryocopus pileatus) ...

## ABSTRACT



## INTRODUCTION
[your paper does not need an introduction]

## METHODS

[your paper does not need to have any field methods]

### STATISTICAL METHODS 














## RESULTS


There was no significant relationship between the number of Pileated woodpeckers obseved on USGS routes in PA and time (F = 0.05, p = 0.95)














## DISCUSSION
[your paper does not need a discussion]

## AKNOWLEDGEMENT
[skip]


## FIGURES

### Figure 1
```{r echo = FALSE}
ggerrorplot(data = PIWO_00_05_10,
            y = "SpeciesTotal",
            x = "Year",
            desc_stat = "mean_ci",
            xlab = "Year",
            ylab = "Number of PIWO observed",
            ylim = c(0, 1.5))

```


### Figure 1 caption
















## APPENDICES

### Appendix 1: Data dictionary
[Paste your data dictionary as a table into your paper]

### Appendix 2: Exploratory graphs

#### Exploratory graphs: boxplots

```{r echo = F}
ggboxplot(data = PIWO_00_05_10,
          y = "SpeciesTotal",
          x = "Year")

```

caption














#### Exploratory graphs: histograms
```{r echo = F}
gghistogram(data = PIWO_00_05_10,
            x = "SpeciesTotal",
            facet.by = "Year")
```

caption











## SUPLEMENTAL FILES

### Supplemental File 1: Raw data file: 
Raw data is contained in Brouwer_NL_PIWO_vs_forest_cover.csv


### Supplemental file 2: Data analysis script
All code to reproduce these analyses is in the script file BROUWER_NL_analysis_script_PIWO_vs_forest_cover.R

