---
title: "ENS 495 Fall 2017 Independent Data Analysis: Example Outline"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


```{r echo = FALSE}
##source analysis script
 source("Brouwer_NL_analysis_script_PIWO_vs_forest_cover.R", encoding = 'UTF-8')
```

## TITLE: Relationship between Pileated Woodpecker (Dryocopus pileatus) abundance and forest cover in Pennsylvania: An analysis using USGS Breeding Bird Survey data

## ABSTRACT

I found a significant positive relationshp between Pileated Woodpecker abundance on BBS routes and both mixed and deciduous forest cover.  The relationship was mixed for, however, was much steeper and had a better fit to the data.

## INTRODUCTION
[your paper does not need an introduction]

## METHODS

[your paper does not need to have any field methods]

### STATISTICAL METHODS 

The relationship between forest cover and the abundance of Pileated Woodpecker on each BBS route was modeled using linear regression in R (R Core Team 2017).  Modeles were evaluated using hypothesis tests and AIC.

## RESULTS

### Hypothesis testing

There was a significant relationship between the number of Pileated Woodpeckers observed on a BBS route and both the percentage of deciduous forest cover (F = 10.61, P = 0.0014) and mixed forest cover (F = 44.55, P < 0.0001).  Both relationships were positive but the slope of the mixed forest regression was much steeper (slope = 16.12, SE = 2.4) than the deciduous forest regression (slope = 1.58, SE = 0.48).  Moreover, the mixed forest regression had a larger R^2 than the deciduous forest model (mixed forest R^2 = 0.24; deciduous forest R^2 = 0.066) and also a much lower AIC value (deciduous versus mixed delta AIC = 28.7)


## DISCUSSION
[your paper does not need a discussion]

## AKNOWLEDGEMENT
[skip]


## FIGURES

### Figure 1
```{r echo = FALSE}
ggscatter(data = BBS_PA_PIWO_4,
          y = "SpeciesTotal",
          x = "mixed.forest.percent",
          add = "reg.line",
          conf.int = TRUE,
          xlab = "Percent mixed forest landcover",
          ylab = "Number of PIWO observed on BBS route")
```


### Figure 1 caption
Relationship between the percent of mixed forest cover around a BBS route and the number of Pileated Woodpeckers (Dryocopus pileatus) observed.  Error band = 95% confidence interval.

## APPENDICES

### Appendix 1: Data dictionary
[Paste your data dictionary as a table into your paper]

### Appendix 2: Exploratory graphs

#### Exploratory graphs: boxplots

```{r echo = F}
plot_grid(plot1.spp.total,
          plot2.decid.percent,
          plot3.confir.percent,
          plot4.mixed.percent,
          labels = c("A", "B",
                      "C","D"))
```
Boxplots showing the distribution of the abundance of Pileated Woodpeckers on each BBS route (A) and  the distribution of percent deciduous forest cover (B), confir forest cover (C) and mixed forest cover.  Deciduous and mixed forest cover were modeld; confir is shown only for compariosns

#### Exploratory graphs: histograms
```{r echo = F}
plot_grid(plot1.spp.total,
          plot2.decid.percent,
          plot3.confir.percent,
          plot4.mixed.percent,
          labels = c("A", "B",
                     "C","D"))
```
Histograms showing the distribution of the abundance of Pileated Woodpeckers on each BBS route (A) and  the distribution of percent deciduous forest cover (B), confir forest cover (C) and mixed forest cover.  Deciduous and mixed forest cover were modeld; confir is shown only for compariosns


## SUPLEMENTAL FILES

### Supplemental File 1: Raw data file: 
Raw data is contained in Brouwer_NL_PIWO_vs_forest_cover.csv


### Supplemental file 2: Data analysis script
All code to reproduce these analyses is in the script file BROUWER_NL_analysis_script_PIWO_vs_forest_cover.R

